---
title: "Using {utVizSankey}"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using {utVizSankey}}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(utVizSankey)
```

## An example dataset

In the `utVizSankey` package, an example dataset `admissions` is available.

```{r load_admissions_data}
data(admissions, package = "utVizSankey")

head(admissions)
```

The dataset contains `r ncol(admissions)` variables and `r nrow(admissions)` observations.

## A simple sankey chart

The function `sankey()` generates a sankey chart. The two main arguments to this function are:

- `data` (which provides the data-frame for the chart);
- and `steps` (indicating which columns of that data-frame should be presented in the chart, from
left-to-right).

```{r basic_sankey}
sankey(
  admissions,
  steps = c("student_type", "gpa", "outcome")
)
```

To alter the initial size of the chart, the `height` and `width` arguments to `sankey()` can be
modified:

```{r sized_sankey}
sankey(
  admissions,
  steps = c("student_type", "gpa", "outcome"),
  height = "100%",
  width = "100%"
)
```

Steps can also be provided as a list containing vectors.
If a given step is a vector, then each element represents a drill-down stage for that column
in the Sankey diagram, starting from the first element. For example, a list of
`list(c("a", "b"), "c", "d")` will create a Sankey with three steps or columns which initially
shows data corresponding to the properties for the keys "a", "c", and "d" in the data.
Clicking on any node in the first column will drill-down to show the data for the "b" key,
filtered on a particular value for the "a" property.

```{r nested_steps}
sankey(
  admissions,
  steps = list(c("student_type", "college"), "gpa", "outcome"),
  width = "100%",
  height = "100%"
)
```

## Sankey pop-ups

When the user hovers over either a node or a link in the Sankey visualization, a pop-up showing
various information about the nodes / links is displayed.

By default:

- the node pop-up shows the name of the node and the count of individuals associated with that node;
- and the link pop-up shows the source and target of the link, the number of individuals associated
with that link and the proportion of individuals associated with source and target that pass through
that link.

The content of these pop-ups can be customized by passing a `node_template` or `link_template`
argument to the `sankey()` function. These strings use the
["Handlebars" templating language](https://handlebarsjs.com/).
Within the template strings, you can reference different properties of the hovered node or link by
including that property in 'double-braces' (`{{someProperty}}`). The properties that are available
for nodes and links are documented in the help page for the function. See `? sankey`.

Here we add a template string for the link pop-up that displays the name of the source and target of
the link and the number of individuals associated with that link. Node templates are constructed
similarly.

```{r}
# Note we use R's raw strings r"(...)" to allow embedding quotes inside the string
link_template = r"("{{sourceName}}" - "{{targetName}}": {{count}} of the {{totalCount}} students)"

sankey(
  admissions,
  steps = list(c("student_type", "college"), "gpa", "outcome"),
  link_template = link_template,
  width = "100%",
  height = "100%"
)
```

## Colors

Colors on the Sankey visualization can be customized by defining the default line colors, the hover
colors, and a list of color overrides on top of these.

### Setting line and hover colors

Starting simple, we can customize the line color and the hover color when we create the Sankey
diagram. We specify the colors in any valid format that would be used in R, such as

- An [R color name](https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/colors.html),
such as `"black"`, `"magenta"`, or `darkslateblue"`.
- A hex code representing the color. For example, the R color "hotpink" can be specified as as the
string `"#FF69B4"`. Letters in the hex code are not case sensitive.
- An rgb specification. For example, the R color "forestgreen" can be specified as
`"rgb(34, 139, 34, maxColorValue = 255)"`

For example, if we choose the black and orange from the Okabe-Ito colorblind-safe palette as the
base and hover colors, this would be a valid method for specifying the colors:

```{r sankey_with_colors}
sankey(
  admissions,
  steps = list(c("student_type", "college"), "gpa", "outcome"),
  color = "black",
  hover_color = "#e69f00"
)
```

When choosing colors, note that some transparency will be applied to the colors to help with the
viewing of overlapping lines.

### Color overrides

You can manually override the colors for particular groups or categories within a group by
performing color overrides. Overrides are provided as a list, where each override is a list
containing at least a color and a group, and if you want to specify that the color should only be
applied to one category in that group, you can specify the name of the category.

```{r color_overrides}
sankey(
  admissions,
  steps = list(c("student_type", "college"), "gpa", "outcome"),
  color_overrides = list(
    list(color = "#009e73", group = "gpa", name = "high"),
    list(color = "#cc79a7", group = "student_type")
  )
)
```

Note that when clicking on a student type in the groups on the left, the color override does not
apply to the subgroup, but a separate color override can be specified on the subgroup if it is
needed.

## Using {utVizSankey} in a shiny app

Example shiny apps that use `utVizSankey` are stored within the package. The source code can be
found in `./inst/shiny-examples/<app-name>`. To run these type
`run_shiny_example(app_name = "the-app")`.

Advanced: Data can be obtained from a node in a Sankey chart by alt-clicking on the node. A shiny
input-handler `utVizSankey.nodeConverter` is provided that will ensure the data is a data.frame when
it reaches R. To bind that data.frame to a shiny 'input' variable, you must provide an
`alt_click_handler` argument to the `sankey()` function. This should be a JavaScript function
created with `htmlwidgets::JS()` that takes the arguments "event" and "data".

```r
alt_click_handler = htmlwidgets::JS(r"(function(event, data){
  Shiny.setInputValue("a_shiny_input_var:utVizSankey.nodeConverter", data.entries)
})")

sankey(data, steps, alt_click_handler = alt_click_handler)
```

This binds the data associated with a Sankey node to the shiny variable `input$a_shiny_input_var`.
